---
title: "GMO Safety Statistical Analysis"
author: "Nazrin Shahpalangova"
date: "2026-06-15"
output: html_document
---

## Introduction

This report presents a statistical analysis of GMO product safety data, including risk score distributions and ANOVA tests across crop categories.

## Load Data

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggplot2)
library(dplyr)

df <- read_csv("C:/Users/SARN/gmo-safety-dashboard/data/gmo_dataset_clean.csv")
head(df)
```

## Dataset Overview

```{r overview}
cat("Total GMO products:", nrow(df), "\n")
cat("Crop types:", length(unique(df$crop_type)), "\n")
summary(df$risk_score)
```

## ANOVA Analysis

```{r anova}
anova_result <- aov(risk_score ~ crop_type, data = df)
summary(anova_result)
```

**Result:** p = 3.63e-05 — Risk scores differ significantly across crop types.

## Risk Score by Crop Type

```{r boxplot, fig.width=10, fig.height=6}
ggplot(df, aes(x = crop_type, y = risk_score, fill = crop_type)) +
  geom_boxplot() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none") +
  labs(title = "Risk Score Distribution by Crop Type",
       x = "Crop Type", y = "Risk Score")
```

## Approval Timeline

```{r scatter, fig.width=10, fig.height=6}
ggplot(df, aes(x = approval_year, y = risk_score, color = crop_type)) +
  geom_point(size = 3) +
  theme_minimal() +
  labs(title = "Approval Year vs Risk Score",
       x = "Approval Year", y = "Risk Score")
```

## Conclusion

ANOVA analysis confirms statistically significant differences in risk scores across GMO crop categories (p < 0.001). Animal-based GMOs show the highest risk scores, while Apple and Potato varieties show the lowest.